---
title: "Portland metropolitan area COVID19 dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    source_code: https://github.com/raford/pdx-COVID19-Dashboard
    social: ["twitter", "facebook", "linkedin"]
runtime: shiny
---

```{r requiredPackages, include=FALSE}
library(flexdashboard)
library(shiny)
library(rmarkdown)
library(plotly)
```

```{r neededFunctions}
# Compute new values
get.new.vals <- function(cum.vec) {
  # This function will compute new values from one unit of time when given a
  # numeric vector.
  # ----------
  # INPUTS
  # cum.vec := a numeric vector containing cummulative values.
  # ----------
  # OUTPUT
  # A vector containing the differenced values from time k+1 and time k, where
  # k is an integer of unit time. NOTE: It is assumed that the difference
  # between time 2 and time 1 is zero. This assumption may not be valid for all
  # types of analyses.
  ind.vals <- diff(cum.vec)
  return(c(cum.vec[1], ind.vals))
}

# Simple moving average function
sma <- function(x, period) {
  # This function will compute a simple moving average of a period equal to
  # period.
  # ----------
  # INPUTS
  # x      := a numeric vector.
  # period := a numerical value that specifies the time period to be used for
  #           the moving average.
  # ----------
  # OUPUT
  # A vector containing the moving average values. Values for which a moving 
  # average could not be computed will have an NA placed in their position.
  dat.vec <- x
  back.len <- period
  vals <- c(rep(x=NA, times=back.len-1))
  for (i in back.len:length(dat.vec)) {
    start <- i-(back.len-1) # Start index
    end <- i # End index
    #print(start:end) # Okay
    #print(length(start:end)) # Okay
    #print(dat.vec[start:end])  
    value <- sum(dat.vec[start:end]) / back.len
    vals <- c(vals, value)
  }
  return(vals)
}

# County data reformatting function
get.county.data <- function(df, county.name, state.name) {
  # This function will create a subset of the data to be used based on the
  # specified County and State and reformat that subset into a more desireable
  # format. We remove the "fips" column as we do not need this data.
  # ----------
  # INPUTS
  # df          := a data frame containing both state and county level data.
  # county.name := the desired county name. Must be a string.
  # state.name  := the desired state name. Must be a string.
  # ----------
  # OUTPUT
  # A dataframe containing the data for the county and state provided as an
  # argument.
  sub.set <- subset(df, county==county.name & state==state.name)[, -4]
  rownames(sub.set) <- NULL
  names(sub.set) <- c("Date", "County", "State", "TotalCases", "TotalDeaths")
  sub.set$NewCases <- get.new.vals(sub.set$TotalCases)
  sub.set$NewDeaths <- get.new.vals(sub.set$TotalDeaths)
  sub.set$Avg7DayNewCases <- sma(sub.set$NewCases, 7)
  sub.set$Avg7DayNewDeaths <- sma(sub.set$NewDeaths, 7)
  return(sub.set)
}

make.plot <- function(df, xvals, yvals, y.sma, Title, max.y) {
  #New cases/deaths
  the.plot <- ggplot(df, aes(x=xvals, y=yvals)) +
  geom_area(aes(color = "orange", fill = "orange"),
            alpha = 0.5, position = position_dodge(0.5)) +
  coord_cartesian(ylim = c(0, max.y)) +
  scale_x_date(date_labels = "%Y %b %d") +
  scale_color_manual(values = c("orange")) +
  scale_fill_manual(values = c("orange")) +
  theme(axis.title.x=element_blank(), axis.title.y = element_blank(),
        legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  geom_line(aes(x=xvals, y=y.sma), color="blue", size=0.75) +
  annotate("text", x=as.Date(mult.dat$Date[50]), y=max.y/1.1, label=Title)
  return(the.plot)
}

cum.plot <- function(df, xvals, yvals, Title, max.y) {
  # Cumulative values plot
  the.plot <- ggplot(df, aes(x=xvals, y=yvals)) +
  geom_area(aes(color = "orange", fill = "orange"),
            alpha = 0.5, position = position_dodge(0.5)) +
  coord_cartesian(ylim = c(0, max.y)) +
  scale_x_date(date_labels = "%Y %b %d") +
  scale_color_manual(values = c("orange")) +
  scale_fill_manual(values = c("orange")) +
  theme(axis.title.x=element_blank(), axis.title.y = element_blank(),
        legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  annotate("text", x=as.Date(mult.dat$Date[50]), y=max.y/1.1, label=Title)
  return(the.plot)
}
```

```{r makeData}

dat <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv")

mult.dat <- get.county.data(dat, county.name="Multnomah", state.name="Oregon")
clack.dat <- get.county.data(dat, county.name="Clackamas", state.name="Oregon")
wash.dat <- get.county.data(dat, county.name="Washington", state.name="Oregon")
clark.dat <- get.county.data(dat, county.name="Clark", state.name="Washington")

rm(dat)
```

Cases
=====================================
Row {.tabset .tabset-fade}
-------------------------------------
   
### New Cases by County (with 7-day Moving Average)

```{r}
max.t.new.cases <- max(mult.dat$NewCases, wash.dat$NewCases,
                       clark.dat$NewCases, clack.dat$NewCases)

mult.new.cases <- make.plot(df=mult.dat, xvals=as.Date(mult.dat$Date),
                    yvals=mult.dat$NewCases, y.sma=mult.dat$Avg7DayNewCases,
                    Title="Multnomah County, OR", max.y=max.t.new.cases)

wash.new.cases <- make.plot(df=wash.dat, xvals=as.Date(wash.dat$Date),
                    yvals=wash.dat$NewCases, y.sma=wash.dat$Avg7DayNewCases,
                    Title="Washington County, OR", max.y=max.t.new.cases)

clark.new.cases <- make.plot(df=clark.dat, xvals=as.Date(clark.dat$Date),
                    yvals=clark.dat$NewCases, y.sma=clark.dat$Avg7DayNewCases,
                    Title="Clark County, WA", max.y=max.t.new.cases)

clack.new.cases <- make.plot(df=clack.dat, xvals=as.Date(clack.dat$Date),
                    yvals=clack.dat$NewCases, y.sma=clack.dat$Avg7DayNewCases,
                    Title="Clackamas County, OR", max.y=max.t.new.cases)

subplot(mult.new.cases, wash.new.cases, clark.new.cases, clack.new.cases, 
        nrows=2, shareX=TRUE, shareY=TRUE)
```   

### Cumulative Cases by County
```{r}
max.t.cum.cases <- max(mult.dat$TotalCases, wash.dat$TotalCases,
                       clark.dat$TotalCases, clack.dat$TotalCases)

mult.cum.cases <- cum.plot(df=mult.dat, xvals=as.Date(mult.dat$Date),
                    yvals=mult.dat$TotalCases, Title="Multnomah County, OR",
                    max.y=max.t.cum.cases)

wash.cum.cases <- cum.plot(df=wash.dat, xvals=as.Date(wash.dat$Date),
                    yvals=wash.dat$TotalCases, Title="Washington County, OR",
                    max.y=max.t.cum.cases)

clark.cum.cases <- cum.plot(df=clark.dat, xvals=as.Date(clark.dat$Date),
                    yvals=clark.dat$TotalCases, Title="Clark County, WA",
                    max.y=max.t.cum.cases)

clack.cum.cases <- cum.plot(df=clack.dat, xvals=as.Date(clack.dat$Date),
                    yvals=clack.dat$TotalCases, Title="Clackamas County, OR",
                    max.y=max.t.cum.cases)

subplot(mult.cum.cases, wash.cum.cases, clark.cum.cases, clack.cum.cases, 
        nrows=2, shareX=TRUE, shareY=TRUE)
```


Deaths
=====================================
Row {.tabset .tabset-fade}
-------------------------------------
   
### New Deaths by County (with 7-day Moving Average)

```{r}
max.t.new.deaths <- max(mult.dat$NewDeaths, wash.dat$NewDeaths,
                       clark.dat$NewDeaths, clack.dat$NewDeaths)

mult.new.deaths <- make.plot(df=mult.dat, xvals=as.Date(mult.dat$Date),
                    yvals=mult.dat$NewDeaths, y.sma=mult.dat$Avg7DayNewDeaths,
                    Title="Multnomah County, OR", max.y=max.t.new.deaths)

wash.new.deaths <- make.plot(df=wash.dat, xvals=as.Date(wash.dat$Date),
                    yvals=wash.dat$NewDeaths, y.sma=wash.dat$Avg7DayNewDeaths,
                    Title="Washington County, OR", max.y=max.t.new.deaths)

clark.new.deaths <- make.plot(df=clark.dat, xvals=as.Date(clark.dat$Date),
                    yvals=clark.dat$NewDeaths, y.sma=clark.dat$Avg7DayNewDeaths,
                    Title="Clark County, WA", max.y=max.t.new.deaths)

clack.new.deaths <- make.plot(df=clack.dat, xvals=as.Date(clack.dat$Date),
                    yvals=clack.dat$NewDeaths, y.sma=clack.dat$Avg7DayNewDeaths,
                    Title="Clackamas County, OR", max.y=max.t.new.deaths)

subplot(mult.new.deaths, wash.new.deaths, clark.new.deaths, clack.new.deaths, 
        nrows=2, shareX=TRUE, shareY=TRUE)
```   

### Cumulative Deaths
```{r}
max.t.cum.deaths <- max(mult.dat$TotalDeaths, wash.dat$TotalDeaths,
                       clark.dat$TotalDeaths, clack.dat$TotalDeaths)

mult.cum.deaths <- cum.plot(df=mult.dat, xvals=as.Date(mult.dat$Date),
                    yvals=mult.dat$TotalDeaths, Title="Multnomah County, OR",
                    max.y=max.t.cum.deaths)

wash.cum.deaths <- cum.plot(df=wash.dat, xvals=as.Date(wash.dat$Date),
                    yvals=wash.dat$TotalDeaths, Title="Washington County, OR",
                    max.y=max.t.cum.deaths)

clark.cum.deaths <- cum.plot(df=clark.dat, xvals=as.Date(clark.dat$Date),
                    yvals=clark.dat$TotalDeaths, Title="Clark County, WA",
                    max.y=max.t.cum.deaths)

clack.cum.deaths <- cum.plot(df=clack.dat, xvals=as.Date(clack.dat$Date),
                    yvals=clack.dat$TotalDeaths, Title="Clackamas County, OR",
                    max.y=max.t.cum.deaths)

subplot(mult.cum.deaths, wash.cum.deaths, clark.cum.deaths, clack.cum.deaths, 
        nrows=2, shareX=TRUE, shareY=TRUE)
```

Data
=====================================
Row {.tabset .tabset-fade}
-------------------------------------

### Multnomah County, OR
```{r}
renderTable(mult.dat)
```

### Washington County, OR
```{r}
renderTable(wash.dat)
```

### Clark County, WA
```{r}
renderTable(clark.dat)
```

### Clackamas County, OR
```{r}
renderTable(clack.dat)
```


About
=====================================
This dashboard is currently a work-in-progress.

I hope to have this About section updated no later than 26 August 2020.

Data: Data comes from the NYT COVID data respository on GitHub: https://github.com/nytimes/covid-19-data

Questions/comments/concerns: Email me at raymond.anthony.ford@gmail.com